Endogenous Macrodynamics in Algorithmic Recourse

### 2023-01-17 intro.qmd
sorted out charts

N = 1000
xmax = 2
X, ys = make_blobs(
    N, 2; 
    centers=2, as_table=false, center_box=(-xmax => xmax), cluster_std=0.1
)
ys .= ys.==2
X = X'
xs = Flux.unstack(X,2)
data = zip(xs,ys)
counterfactual_data = CounterfactualData(X,ys')
plot()
scatter!(counterfactual_data)
```

@@

```{julia}
n_epochs = 100
model = Chain(Dense(2,1))
mod = FluxModel(model)
Models.train(mod, counterfactual_data; n_epochs=n_epochs)
```

@@

```{julia}
opt = Flux.Adam(0.01)
gen = GenericGenerator(;decision_threshold=γ, opt=opt)
```

@@

plt = plot(plt_original, plt_single, plt_single_retrained, plt_single_repeat, layout=(1,4), legend=false, axis=nothing, size=(600,165))
savefig(plt, joinpath(www_path, "poc.png"))
display(plt)
```

>>>>

max_obs = 1000
catalogue = load_synthetic(max_obs)
counterfactual_data = catalogue[:linearly_separable]
X = counterfactual_data.X
ys = vec(counterfactual_data.y)
plot()
scatter!(counterfactual_data)
```

@@

```{julia}
n_epochs = 100
model = Chain(Dense(2,1))
mod = FluxModel(model)
Models.train(mod, counterfactual_data; n_epochs=n_epochs)
mod_orig = deepcopy(mod)
```

@@

```{julia}
opt = Flux.Descent(0.01)
gen = GenericGenerator(;decision_threshold=γ, opt=opt)
```

@@

plt = plot(plt_original, plt_single, plt_single_retrained, plt_single_repeat, layout=(1,4), legend=false, axis=nothing, size=(600,165))
savefig(plt, joinpath(www_path, "poc.png"))
savefig(plt, "paper/www/poc.png")
display(plt)
```

## Mitigation Strategies

```{julia}
#| output: true
#| label: fig-mitigate
#| fig-cap: "Mitigation strategies."

# Generators:
generators = Dict(
    "Generic (γ=0.5)" => GenericGenerator(opt = opt, decision_threshold=0.5),
    "Generic (γ=0.9)" => GenericGenerator(opt = opt, decision_threshold=0.9),
    "Gravitational" => GravitationalGenerator(opt = opt),
    "ClaPROAR" => ClapROARGenerator(opt = opt)
)

# Counterfactuals
x = select_factual(counterfactual_data, rand(candidates)) 
counterfactuals = Dict([name => generate_counterfactual(x, 1, counterfactual_data, mod_orig, gen;) for (name, gen) in generators])

# Plots:
plts = []
for (name,ce) ∈ counterfactuals
    plt = plot(ce; title=name, colorbar=false, ticks = false, legend=false, zoom=0)
    plts = vcat(plts..., plt)
end
plt = plot(plts..., size=(750,200), layout=(1,4))
savefig(plt, joinpath(www_path, "mitigation.png"))
savefig(plt, "paper/www/mitigation.png")
display(plt)
```

### 2023-01-17 paper/paper.tex
sorted out charts

\caption{Dynamics in Algorithmic Recourse: (a) we have a simple linear classifier trained for binary classification where samples from the negative class ($y=0$) are marked in blue and samples of the positive class ($y=1$) are marked in orange; (b) the implementation of AR for a random subset of individuals leads to a noticable domain shift; (c) as the classifier is retrained we observe a corresponding model shift; (d) as this process is repeated, the decision boundary moves away from the target class.}\label{fig:poc}
\end{figure}

We think that these types of endogenous dynamics may be problematic and deserve our attention. From a purely technical perspective we note the following: firstly, model shifts may inadvertently change classification outcomes for individuals who never received and implemented recourse. Secondly, we observe in Figure \ref{fig:poc} that as the decision boundary moves in the direction of the non-target class, counterfactual paths become shorter. We think that in some practical applications, this can be expected to generate costs for involved stakeholders. To follow our argument, consider the following two examples:

\begin{example}[Consumer Credit]
\protect\hypertarget{exm:consumer}{}\label{exm:consumer}Suppose Figure \ref{fig:poc} relates to an automated decision-making system used by a retail bank to evaluate credit applicants with respect to their creditworthiness. Assume that the two features are meaningful in the sense that creditworthiness increases in the south-east direction. Then we can think of the outcome in panel (d) as representing a situation where the bank supplies credit to more borrowers (orange), but these borrowers are on average less creditworthy and more of them can be expected to default on their loan. This represents a cost to the retail bank.
\end{example}

\begin{example}[Student Admission]
\protect\hypertarget{exm:student}{}\label{exm:student}Suppose Figure \ref{fig:poc} relates to an automated decision-making system used by a university in its student admission process. Assume that the two features are meaningful in the sense that the likelihood of students completing their degree increases in the south-east direction. Then we can think of the outcome in panel (b) as representing a situation where more students are admitted to university (orange), but they are more likely to fail their degree than students that were admitted in previous years. The university admission committee catches on to this and suspends its efforts to offer Algorithmic Recourse. This represents an opportunity cost to future student applicants, that may have derived utility from being offered recourse.
\end{example}

@@

where \(X=\{x_1,...,x_m\}\), \(\tilde{X}=\{\tilde{x}_1,...,\tilde{x}_n\}\) represent independent and identically distributed samples drawn from probability distributions \(\mathcal{X}\) and \(\mathcal{\tilde{X}}\) respectively \protect\hyperlink{ref-gretton2012kernel}{{[}25{]}}. MMD is a measure of the distance between the kernel mean embeddings of \(\mathcal{X}\) and \(\mathcal{\tilde{X}}\) in a Reproducing Kernel Hilbert Space, \(\mathcal{H}\) \protect\hyperlink{ref-berlinet2011reproducing}{{[}26{]}}. An important consideration is the choice of the kernel function \(k(\cdot,\cdot)\). In our implementation we make use of a Gaussian kernel with a constant length-scale parameter of \(0.5\). As the Gaussian kernel captures all moments of distributions \(\mathcal{X}\) and \(\mathcal{\tilde{X}}\), we have that \(MMD(X,\tilde{X})=0\) if and only if \(X=\tilde{X}\). Conversely, larger values \(MMD(X,\tilde{X})>0\) indicate that it is more likely that \(\mathcal{X}\) and \(\mathcal{\tilde{X}}\) are different distributions. In our context, large values therefore indicate that a domain shift indeed seems to have occurred.

>>>>

\caption{Dynamics in Algorithmic Recourse: (a) we have a simple linear classifier trained for binary classification where samples from the negative class ($y=0$) are marked in orange and samples of the positive class ($y=1$) are marked in blue; (b) the implementation of AR for a random subset of individuals leads to a noticeable domain shift; (c) as the classifier is retrained we observe a corresponding model shift; (d) as this process is repeated, the decision boundary moves away from the target class.}\label{fig:poc}
\end{figure}

We think that these types of endogenous dynamics may be problematic and deserve our attention. From a purely technical perspective, we note the following: firstly, model shifts may inadvertently change classification outcomes for individuals who never received and implemented recourse. Secondly, we observe in Figure \ref{fig:poc} that as the decision boundary moves in the direction of the non-target class, counterfactual paths become shorter. We think that in some practical applications, this can be expected to generate costs for involved stakeholders. To follow our argument, consider the following two examples:

\begin{example}[Consumer Credit]
\protect\hypertarget{exm:consumer}{}\label{exm:consumer}Suppose Figure \ref{fig:poc} relates to an automated decision-making system used by a retail bank to evaluate credit applicants with respect to their creditworthiness. Assume that the two features are meaningful in the sense that creditworthiness increases in the South-East direction. Then we can think of the outcome in panel (d) as representing a situation where the bank supplies credit to more borrowers (orange), but these borrowers are on average less creditworthy and more of them can be expected to default on their loan. This represents a cost to the retail bank.
\end{example}

\begin{example}[Student Admission]
\protect\hypertarget{exm:student}{}\label{exm:student}Suppose Figure \ref{fig:poc} relates to an automated decision-making system used by a university in its student admission process. Assume that the two features are meaningful in the sense that the likelihood of students completing their degree increases in the South-East direction. Then we can think of the outcome in panel (b) as representing a situation where more students are admitted to university (orange), but they are more likely to fail their degree than students that were admitted in previous years. The university admission committee catches on to this and suspends its efforts to offer Algorithmic Recourse. This represents an opportunity cost to future student applicants, that may have derived utility from being offered recourse.
\end{example}

@@

where \(X=\{x_1,...,x_m\}\), \(\tilde{X}=\{\tilde{x}_1,...,\tilde{x}_n\}\) represent independent and identically distributed samples drawn from probability distributions \(\mathcal{X}\) and \(\mathcal{\tilde{X}}\) respectively \protect\hyperlink{ref-gretton2012kernel}{{[}25{]}}. MMD is a measure of the distance between the kernel mean embeddings of \(\mathcal{X}\) and \(\mathcal{\tilde{X}}\) in a Reproducing Kernel Hilbert Space, \(\mathcal{H}\) \protect\hyperlink{ref-berlinet2011reproducing}{{[}26{]}}. An important consideration is the choice of the kernel function \(k(\cdot,\cdot)\). In our implementation, we make use of a Gaussian kernel with a constant length-scale parameter of \(0.5\). As the Gaussian kernel captures all moments of distributions \(\mathcal{X}\) and \(\mathcal{\tilde{X}}\), we have that \(MMD(X,\tilde{X})=0\) if and only if \(X=\tilde{X}\). Conversely, larger values \(MMD(X,\tilde{X})>0\) indicate that it is more likely that \(\mathcal{X}\) and \(\mathcal{\tilde{X}}\) are different distributions. In our context, large values, therefore, indicate that a domain shift indeed seems to have occurred.
